---
title: "Class1"
output: html_document
date: "2025-10-16"
---

### NAME: Grace TETA
### ID: 101378
### CLASS 1
-----------------------------------------------------

### Vector: The foundational R data structure
### Numeric
```{r}
 x <- c(0.5, 0.6)
 age<-c(20,35,32,29)
 summary(age)
```
```{r}
 age<-c(35,24,18,24)
 mean(age)
```
### Logical
```{r}
x <- c(TRUE, FALSE)
 x <- c(T, F)

```

### Character
```{r}
x<-c("a","b","c")
 class<-c("M","F","F","M")
```
### Integer
```{r}
x<-9:29
```
### Complex
```{r}
x<-c(1+0i,2+4i)
```
### Matrixandoperationwithinmatrices
```{r}
 x<-matrix(1:6,nrow=2,ncol=3)
 x
```
```{r}
 (y<-matrix(1:6,nrow=3,ncol=2))
```
```{r}
 x%*%y
```
### List
```{r}
 x<-list(1,"a",TRUE,1+4i)
 x
```
### Factor
```{r}
 x<-factor(c("yes","yes","no","yes","no"))
```
###  Create a vector with NAs in it
```{r}
x<-c(1,2,NA,10,3)
# Return a logical vector indicating which elements are NA
is.na(x)

```
```{r}
x<-c(1,2,4,"NA",5)
 bad<-is.na(x)
 print(bad)
```
```{r}
x[!bad]
```
###  What if there are multiple R objects and you want to take the subset with no missing values in any of those objects?
```{r}
x<-c(1,2,NA,4,NA,5)
 y<-c("a","b",NA,"d",NA,"f")
 good<-complete.cases(x,y)
  good
```
###  Coercion
```{r}
 sum(c(TRUE,TRUE,FALSE,FALSE,FALSE))
```
### Create data
### ###Let’s create a data frame
```{r}
 studentID<-c(1,2,3,4,5)
 math_score<-c(12,17,10,9,NA)
 gender<-c("M","F","M","M","F")
 it_score<-c(13,18,11,10,19)
 scoredata<-data.frame(studentID,gender,math_score,it_score)
 scoredata

```
###  Create data from keyboard
### Steps1.Create data frame (ormatrix) with variable names2.Invoke the text editor in the data objected created at first step
```{r}
data_class2<-data.frame(height=numeric(0),weight=numeric(0),bmi=numeric
 (0))
 data_class2<-edit(data_class2)
```

###  Data Importation
### HW1:Import data from Statistical package and from Database management systems
```{r}
bmi_data <- read.csv("C:\\Users\\USER\\Downloads\\bmi.csv")
variable.names(bmi_data)
```
```{r}
 head(bmi_data)
```
```{r}
tail(bmi_data)
```
```{r}
bmi_data[10:20,]
```
```{r}
 summary(bmi_data$WEIGHT)
```
```{r}
length(bmi_data$WEIGHT)
```
```{r}
 bmi_data[,-1]
```
###  Incase you want to use dataset built in R
```{r}
data()#listofdatasetscurrentlyavailable
 data("airquality")
 variable.names(airquality)
```
```{r}
 str(airquality)
```
### Exploration data analysis
### Examine numerical variable using common summary

```{r}
summary(bmi_data)

```
```{r}
 str(bmi_data)
```
###  Summary Statistics
```{r}
 summary(bmi_data)
```
```{r}
 apply(bmi_data,2,mean)
```
```{r}
bmi_data$Height <- as.numeric(as.character(bmi_data$Height))
bmi_data$Weight <- as.numeric(as.character(bmi_data$Weight))
apply(bmi_data,2,sd)
```
```{r}
 c(mean(bmi_data$Height),sd(bmi_data$Height))
```
```{r}
 c(mean(bmi_data$Weight),sd(bmi_data$Weight))
```
```{r}
c(Mean=mean(bmi_data$Height),SD=sd(bmi_data$Height))
```
```{r}
 c(Mean=mean(bmi_data$Weight),SD=sd(bmi_data$Weight))
```
### Variation and covariation using both numerical and graphs
### Relationship between a categorical and a continuous variable
```{r}
data("iris")
 head(iris)
```
```{r}
mean(iris$Petal.Length)
```
```{r}
 mean(iris$Petal.Length[iris$Species=="setosa"])
```
```{r}
 mean(iris$Petal.Length[iris$Species=="versicolor"])
```
```{r}
 mean(iris$Petal.Length[iris$Species=="virginica"])
```
### By shortcut
```{r}
by(iris$Petal.Length,iris$Species,mean)
```
```{r}
by(iris$Petal.Length,iris$Species,sd)
```
```{r}
by(iris$Petal.Length,iris$Species,summary)
```
###  Visualize the differences in continuous variables between categories using Box-and whisker plot
```{r}
boxplot(iris$Sepal.Length~iris$Species,data=iris, main="Comparison")
```
###  Relationship between two categorical variables
```{r}
 ucba<-data.frame(UCBAdmissions)
 head(ucba)
```
```{r}
cross<-xtabs(Freq~Gender+Admit,data=ucba)
 (cross<-xtabs(Freq~Gender+Admit,data=ucba))
```
### Is there gender bias in UCB graduate admission process?
```{r}
 prop.table(cross,2)
```
### Simpson’sparadox
```{r}
 cross2<-xtabs(Freq~Gender+Admit,data=ucba[ucba$Dept=="A",])
 prop.table(cross2,1)
```
### BarChart
```{r}
 dat<-read.table(text="ProdAProdBProdCProdD
 1110506070
 2120508065",header=TRUE)
 barplot(as.matrix(dat),beside=FALSE,col=c("Red","green"))

```
```{r}
 dat <- read.table(text = "A B C D E F G
 1 10 80 30 90 70 60 90
 2 20 50 70 50 40 10 40
 3 60 80 80 60 60 30 160
 4 20 40 70 80 20 10 70", header = TRUE)
 barplot(as.matrix(dat))

```
### Scatter Plot
```{r}
plot(Weight~Height,data=bmi_data)
```
```{r}
plot(bmi_data$Height,bmi_data$Weight)
```
```{r}
plot(bmi_data[,1],bmi_data[,2])
```
```{r}
cor(bmi_data$Height,bmi_data$Weight)
```
###  Linear Models
```{r}
 plot(Weight~Height,data=bmi_data)
 abline(lm(Weight~Height,data=bmi_data)$coefficient)
```
```{r}
 data_lm<-lm(Weight~Height,data=bmi_data)
 summary(data_lm)
```
```{r}
hist(bmi_data$Height)
```
```{r}
hist(bmi_data$Weight,col = "blue")

```
### Linear with other line
```{r}
 plot(Weight~Height,data=bmi_data)
 abline(lm(Weight~Height,data=bmi_data)$coefficient)
 lines(lowess(bmi_data$Height,bmi_data$Weight),col="blue")

```
```{r}
cor(bmi_data$Height,bmi_data$Weight)
```
```{r}
model1<-lm(Weight~Height,data=bmi_data)
```
### Data management
### Create another Variable
```{r}
 attach(bmi_data)
(BodyComposition<-Weight/(Height/100)^2)
```
```{r}
 (BodyComposition<-round(Weight/(Height/100)^2,digit=2))
```
```{r}
 head(cbind(bmi_data,BodyComposition),n=5)
```
```{r}
new_data_class<-cbind(bmi_data,BodyComposition)
 tail(cbind(bmi_data,BodyComposition),n=10)
```
###  Summary of Body Composition
```{r}
 summary(BodyComposition)
```
```{r}
 stem(BodyComposition,scale=2)
```



