---
title: "trial"
output: html_document
date: "2025-10-12"
---

### title: "Class Assignment"
### Names: "Grace TETA"
### date: "2025-10-12"
### ID: "101378"

### Trace ()
Trace function is useful if you need to track down an error which occurs in a base function. The trace() function accepts the name of the function to be traced, a function or unevaluated evaluated expression to execute,and the line number at which to execute it. Line numbers inside a function are revealed by calling list(body(function))

```{r}
my_function <- function(x, y) {
      result <- x + y
      return(result)
    }

    # Trace 'my_function' to insert a browser call at the second line
    trace(my_function, browser, at = 2)

    my_function(5, 3) # This will open the browser at the specified line

    # Untrace the function when done
    untrace(my_function)
```
### Recover()
The recover() statement is used as an error handler. R allows you to choose which function browser to enter and prints the entire call stack in recover(). After then, the debugging session begins at the chosen location.

```{r error=TRUE}
# Calling recover
options(error = recover)

# Function 1
function_1 <- function(a){
 a + 5
}

# Function 2
function_2 <- function(b) {
 function_1(b)
}

# Calling function
function_2("s")
```

